A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-10-28 LyapDraft.tex
Update LyapDraft.tex

To overcome the lack of intuition for the estimate sequences technique, several authors have recently proposed schemes to achieve acceleration without appealing to estimate sequences~\cite{Fazel, BubeckLeeSingh15, Lessard14}.  One promising general approach to the analysis of acceleration has been to analyze the continuous-time limit of accelerated methods and to show that the stability of the resulting ODEs can be verified via a Lyapunov function~\cite{SuBoydCandes14,Krichene15,Acceleration}.  Unfortunately, none of this previous work provides a clear path of obtaining a discrete time optimization algorithm from a continuous time ODE.  There are a vast number of ways to discretize ODEs, but not all of them give rise to convergent methods.  Indeed, for unconstrained optimization on Euclidean space, Polyak's Heavy Ball method and Nesterov's accelerated gradient descent in the setting where the objective is strongly convex have the same continuous time limit.

In this paper, we propose a bridge between the continuous time limits and discrete time algorithms.  Our method takes, as its primary object, a general Lyapunov function for momentum methods in continuous time.  Through a diverse set of examples, we demonstrate how the proofs of momentum methods can be understood as bounding discretization errors of the Lyapunov function when moving to discrete time.  In particular, we show how the discretization of the associated continuous time ODE needs to be performed in such a way that the Lyapunov function remains valid when transitioning to discrete time.
Using this technique, we provide a clean, direct proof of the convergence of Nesterov's method for strongly convex functions in Euclidean spaces.  We explore how only particular parameterizations of the continuous ODEs lead to discrete time methods.  In doing so, we explain the need for the extragradient step inside Nesterov's method which does not appear in Polyak's method.  
Finally, we show there is an equivalence between estimate sequences and Lyapunov functions.  In continuous and discrete time, estimate sequences can be derived directly from the Lyapunov function and vice versa.  We show that the associated continuous time estimate sequence can also be directly discretized to give standard estimate-sequence based proofs of momentum methods.  This clarifies how Nesterov's ``algebraic trick'' is closely related to invariant reduction proofs that are more common in algorithm analysis.

The paper proceeds as follows.  We first describe the related work in viewing optimization algorithms as discretizations of continuous dynamical systems.  We then introduce the notion of Lyapunov functions that we will use throughout the paper to demonstrate convergence of algorithms.  In particular, we will introduce time-varying Lyapunov functions that not only prove algorithm convergence but additionally verify a \emph{convergence rate}.  
%We review how such Lyapunov analysis has been used to analyze mirror decent, reviewing the concepts of Bregman divergence and mirror maps in the process. 
 We then turn to various momentum-based methods, providing a general Lyapunov analysis for Nesterov's accelerated method, the quasi-monotone subgradient method, the conditional gradient method, and a few novel algorithms.  We pay particular attention to the strongly convex case in Euclidean space, highlighting some unique properties of this setup.  Finally, we describe the connection between estimate sequences and Lyapunov functions and directions for future investigation.

>>>>

To overcome the lack of intuition for the estimate sequences technique, several authors have recently proposed schemes to achieve acceleration without appealing to estimate sequences~\cite{Fazel, BubeckLeeSingh15, Lessard14}.  One promising general approach to the analysis of acceleration has been to analyze the continuous-time limit of accelerated methods and to show that the stability of the resulting ODEs can be verified via a Lyapunov function~\cite{SuBoydCandes14,Krichene15,Acceleration}.  However, these methods stop short of providing principles for deriving a discrete-time optimization algorithm from a continuous-time ODE.  There are many ways to discretize ODEs, but not all of them give rise to convergent methods or to acceleration.  Indeed, for unconstrained optimization on Euclidean spaces in the setting where the objective is strongly convex, Polyak's Heavy Ball method and Nesterov's accelerated gradient descent have the same continuous-time limit.

In this paper, we propose a bridge between continuous-time limits and discrete-time algorithms.  Our method takes, as its primary object of study a general Lyapunov function for momentum methods in continuous time.  Through a diverse set of examples, we demonstrate how the proofs of momentum methods can be understood as bounding discretization errors of the Lyapunov function when moving to discrete time.  In particular, we show how the discretization of the associated continuous time ODE needs to be performed in such a way that the Lyapunov function remains valid when transitioning to discrete time.
Using this technique, we provide a simple, direct proof of the convergence of Nesterov's method for strongly convex functions in Euclidean spaces.  Our proof makes clear why only certain parameterizations of the continuous ODEs lead to valid discrete-time methods.  Moreover, we explain the need for the extragradient step inside Nesterov's method which does not appear in Polyak's method.  
Finally, we show there is an equivalence between estimate sequences and Lyapunov functions.  In continuous and discrete time, estimate sequences can be derived directly from the Lyapunov function and vice versa.  We show that the associated continuous-time estimate sequence can also be directly discretized to give standard estimate-sequence based proofs of momentum methods.  This clarifies how Nesterov's ``algebraic trick'' is closely related to invariant reduction proofs that are common in algorithm analysis.

The paper proceeds as follows.  We first describe the related work in viewing optimization algorithms as discretizations of continuous-time dynamical systems.  We then introduce the Lyapunov functions that we will use throughout the paper to demonstrate convergence of algorithms.  In particular, we will introduce time-varying Lyapunov functions that not only establish algorithm convergence but additionally verify a \emph{convergence rate}.  
%We review how such Lyapunov analysis has been used to analyze mirror decent, reviewing the concepts of Bregman divergence and mirror maps in the process. 
 We then turn to various momentum-based methods, providing a general Lyapunov analysis for Nesterov's accelerated method, the quasi-monotone subgradient method, the conditional gradient method, and a few novel algorithms.  We pay particular attention to the strongly convex case in Euclidean space, highlighting some unique properties of this setup.  Finally, we describe the connection between estimate sequences and Lyapunov functions and discuss directions for future investigation.
